Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-19 paper.tex
progress

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{amssymb}

@@

\begin{mylemma} Properties of $\Lift$:
\begin{enumerate}
  \item $\Lift$ preserves all basic type formers; this is immediate from the definition.
  \item $\Lift$ is functorial, i.e.\ $\Lift\,(p \circ q)\,A = \Lift\,p\,(\Lift\,q\,A)$. This follows
    by induction on $A$, and we make use of the irrelevance of $\blank\!<\!\blank$ in the $\U'$ case.
    \qed
\end{enumerate}
\end{mylemma}

@@

\todo{disambiguate IR and TTGU type notation}

\begin{mydefinition}[Base category]
This is simply the category of sets and functions in $\Set0$, i.e.\ $\Con = \Set0$,
$\Sub\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
\end{mydefinition}

\begin{mydefinition}[Family diagram]
We map $i : \Lvl$ to a family structure as follows.
\begin{alignat*}{5}
  &\Ty_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \Tm_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
\end{alignat*}
Type and term substitution are given by composition with some $\sigma : \Gamma
\to \Delta$ function. Context extension is given by $\Gamma \ext A = (\gamma :
\Gamma) \times \El\,(A\,\gamma)$. Since we define $\blank\!\ext\!\blank$ as a
$\Sigma$-type, the comprehension structure for it follows immediately, as
$\Sigma$ pairing and projections.

Mapping $i$ to a strict inclusion is defined as follows. We
\end{mydefinition}

>>>>

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{amssymb}
\usepackage{bm}

@@

\begin{mylemma}\label{lem:liftprop} Properties of $\Lift$:
\begin{enumerate}
  \item $\Lift$ preserves all basic type formers; this is immediate from the definition.
  \item $\Lift$ is functorial, i.e.\ $\Lift\,(p \circ q)\,A = \Lift\,p\,(\Lift\,q\,A)$. This follows
    by induction on $A$, and we make use of the irrelevance of $\blank\!<\!\blank$ in the $\U'$ case.
    \qed
\end{enumerate}
\end{mylemma}

@@

\begin{notation} To avoid name clashing between components of the model and metatheoretic
definitions, we use \textbf{bold} font to refer to TTGU components.
\end{notation}

\begin{mydefinition}[Base category]
This is simply the category of sets and functions in $\Set0$, i.e.\ $\bm{\Con} = \Set0$,
$\bm{\Sub}\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
\end{mydefinition}

\begin{mydefinition}[Family diagram]
We map $i : \Lvl$ to a family structure as follows.
\begin{alignat*}{5}
  &\bm{\Ty}_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \bm{\Tm}_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
\end{alignat*}
Type and term substitution are given by composition with some $\sigma : \Gamma
\to \Delta$ function. Comprehension structure is given by $\Gamma \bm{\ext} A =
(\gamma : \Gamma) \times \El\,(A\,\gamma)$. Type lifting along $p : i < j$ is
as follows:n
\begin{alignat*}{5}
  & \bm{\Lift}_{i}^{j}p : \bm{\Ty}_i\,\Gamma \to \bm{\Ty}_j\,\Gamma\\
  & \bm{\Lift}_{i}^{j}p\,A = \lambda\,\gamma.\,\Lift_{i}^{j}p\,(A\,\gamma)
\end{alignat*}
Now, two of the strict inclusion equations follow from $\msf{ElLift}$, namely
$(\Gamma \bm{\ext} \bm{\Lift}_{i}^{j}p\,A) = (\Gamma \bm{\ext} A)$ and
$\bm{\Tm}_j\,\Gamma\,(\bm{\Lift}_{i}^{j}p\,A) = \bm{\Tm}_i\,\Gamma\,A$. Thus, we
can just define term lifting as $\bm{\up}_{i}^{j}\!p\,t = t$ and
$\bm{\down}_{i}^{j}\!p\,t = t$. Basic type formers are as follows.
\begin{alignat*}{5}
  & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\Gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
  & \hspace{2em}\bm{\Bool} = \lambda\,\gamma.\,\Bool'
\end{alignat*}
$\bm{\Lift}_{i}^{j}p$ preserves type formers by Lemma \ref{lem:liftprop}. We
define basic term formers using metatheoretic term formers, e.g.\ $\bm{\true}_i
= \lambda\,\gamma.\,\true$ and $(\bm{\lambda}_i\,x.\,t) =
\lambda\,\gamma\,\alpha.\,t\,(\gamma,\,\alpha)$. Note that since semantic term
formers are just external constructors, they do not depend on levels, so
e.g.\ $\bm{\true}_i$ is the same at all $i$. This implies that
$\bm{\up}_{i}^{j}\!p$ preserves term formers as well, so ($\bm{\Lift}_{i}^{j}p$,
$\bm{\up}_{i}^{j}\!p$, $\bm{\down}_{i}^{j}\!p$) is a strict family inclusion.

We define universes as $\bm{\U}\,i\,j\,p = \lambda\,\gamma.\,\U'_i\,j\,p$.  With
this, $\bm{\Lift}_{j}^{k}\,p\,(\bm{\U}\,i\,j\,q) = \bm{\U}\,i\,k\,(p \circ q)$
follows by the definition of semantic $\Lift$. The Russell universe equation
$\bm{\Tm}_j\,\Gamma\,(\bm{\U}\,i\,j\,p) = \bm{\Ty}_i\,\Gamma$ follows from Lemma
\ref{lem:ucomp}, so we can define $\bm{\El}$ and $\bm{\Code}$ as identity functions.

\begin{theorem}[Consistency of TTGU]
There is no closed syntactic term of $\bot_i$ for any $i$.
\end{theorem}
\begin{proof}
Assuming a syntactic $t : \Tm_i\,\emptycon\,\bot_i$, we can interpret it in the
previously given model, which yields an inhabitant of the metatheoretic $\bot$,
hence a contradiction.
\end{proof}

%% To verify that we have strict inclusions, we need to
%% show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
%% formers, but first we look at the semantics of type and term formers.

%% To verify that we have strict inclusions, it
%% remains to show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
%% formers.  For $\bm{\Lift}$ this follows from Lemma \ref{lem:liftprop}. For
%% $\bm{\up}$, note that semantics terms

  %% & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
  %% & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
  %% & \up\!t &&= t                                    \label{eq:cumlift}

%% The strict inclusions are given as follows.
  %% Lift : ∀ {Γ i j} → i < j → Ty Γ i → Ty Γ j
  %% Lift p A γ = ↑ p (A γ)

\end{mydefinition}
